### 2018-10-16 Manuscript/Manuscript.tex
Updates from ShareLaTeX

%\usepackage{fontspec}
%\usepackage[utf8]{inputenc}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[normalem]{ulem}
\usepackage[left]{lineno}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{color}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[hidelinks]{hyperref}
\urlstyle{same}
\usepackage{threeparttable}
\doublespacing

@@

The proposed environmental impacts of large igneous province (LIP) eruptions are myriad (as summarized in \citealp{Ernst2017a}). One aspect of LIP emplacement that has been hypothesized to relate to long-term global climate is the effect such emplacement could have on increasing global weatherability and driving cooling. Global weatherability is the sum of factors aside from climate itself that contribute to overall global weathering. On a planet with high weatherability, the CO$_2$ input from volcanism can be removed via silicate weathering at a lower atmospheric CO$_2$ concentration than on a less weatherable planet. The relatively high concentrations of Ca and Mg (that ultimately sequester carbon through precipitation as carbonate) and the high weathering rates of mafic lithologies make it so that basaltic regions consume more CO$_2$ than regions where bedrock composition is closer to bulk continental crust \citep{Dessert2003a}. Data from basaltic watersheds show that chemical weathering rates are greatest in regions with high runoff, and as a result CO$_2$ consumption in basaltic regions is most pronounced in the tropical rain belt \citep{Dessert2003a, Hartmann2014a}.

@@

In the original models that proposed the ``Fire and Ice'' hypothesis as an explanation for the onset of Sturtian glaciation, chemical weathering was modeled as an Arhennius relationship where it was solely a function of temperature and run-off \citep{Donnadieu2004a}. However, such an approach neglects the effects that can result through soil shielding and regolith development in low-relief regions. Recent progress on understanding the relationships between landscapes, topography and chemical weathering reveals that these effects are quite important (e.g. \citealp{Maher2014a}). As a result, more recent modeling of chemical weathering that has incorporated such processes highlights the importance of high-relief regions relative to low-relief ones \citep{Godderis2017b}. Without active uplift, soil shielding from regolith development on low-relief LIPs could significantly decrease the local weatherability of a LIP and mute its impact on global weatherability. In contrast, processes that lead to continued exhumation of mafic lithologies and the creation of steep topography, particularly in tropical regions, may exert a strong control on global weatherability and long-term climate. This interpretation underlies the hypothesis that arc-continent collisons in the tropics during the Ordovician \citep{Swanson-Hysell2017a} and the Cenozoic \citep{Jagoutz2016a} played a significant role in transitions into glacial climate states at those times.

A complication with the interpretation of soil shielding and limited weathering of LIPs is the rapid area decay rate inferred from the comparison of current LIP extent to estimated original extent. A couple considerations are relevant with respect to this analysis: 1) the current extent of LIP exposure is reduced in part by volcanics being covered by unconsolidated sediments (i.e. regolith development itself) in a number of the provinces; 2) the initial LIP extents are typically poorly constrained and are likely overestimates which could be resulting in inflated interpreted decay rates. Future efforts to develop better-constrained estimates of original LIP extent will improve analyses such as that in this contribution.

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$500 Myr. Furthermore, the approach of seeking to evaluate correlation between LIP area and glaciation is complicated for Neoproterozoic pan-glacial events wherein ice-albedo runaway leads to persistent global glaciation absent of continued forcing through normal carbon cycle processes until the build-up of sufficient pCO$_2$ to drive deglaciation. Nevertheless, the  hypothesis of tropical LIP area associated with the ca. 720 Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth glaciation is a major motivating driver behind conducting this analysis. How does the tropical LIP area associated with the Franklin LIP compare to that observed in the Phanerozoic? DISCUSS FRANKLIN AND UMKONDO AREAS AND COMPARE TO THE PHANEROZOIC RECORD.

@@

\begin{table}[h!]
\scriptsize
\caption{place holder}
\noindent\makebox[\textwidth]{
\begin{tabular}{c c c c c}
\hline
\textbf{Fraction} & \textbf{$^{206}$Pb/$^{238}$U} & & \textbf{$^{207}$Pb/$^{235}$U} & \\
& \textbf{Age [Ma]} & $\pm$2$\sigma$ & \textbf{Age [Ma]} & $\pm$2$\sigma$ \\
\hline
z1 & 823.2201669 & 1.067507326 & 823.4322744 & 2.609783978 \\
z2 & 822.401975 & 0.714471249 & 819.0720505 & 1.260737 \\
z4 & 836.2094848& 1.478314118 & 828.1995093 & 6.104877613 \\
z5 & 834.109136 & 1.288954576 & 825.9918302 & 5.720788585 \\
z16 & 831.1612867 & 0.741238812 & 832.226412 & 1.64433912 \\
z17 & 854.9767495 & 0.72532716 & 855.8348689 & 1.16873047 \\
z18 & 812.4046191 & 0.665110984 & 818.3985038 & 1.082055988 \\
z19 & 832.8963617 & 0.694437022 & 833.1583004 & 1.060745556 \\
\hline
\end{tabular}
}
\begin{tablenotes}[para,flushleft]
place holder.
\end{tablenotes}
\label{tab:LIPs}
\end{table}

>>>>

%\usepackage{fontspec}
%\usepackage[utf8]{inputenc}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[normalem]{ulem}
\usepackage[left]{lineno}
\usepackage{booktabs}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{color}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[hidelinks]{hyperref}
\urlstyle{same}
\usepackage{threeparttable}
\doublespacing

@@

The proposed environmental impacts of large igneous province (LIP) eruptions are myriad (as summarized in \citealp{Ernst2017a}). One aspect of LIP emplacement that has been hypothesized to relate to long-term global climate is the effect such emplacement could have on increasing global weatherability and driving cooling. Global weatherability is the sum of factors aside from climate itself, such as the latitudinal distribution of continents and mountain belts, that contribute to overall global weathering. On a planet with high weatherability, the CO$_2$ input from volcanism can be removed via silicate weathering at a lower atmospheric CO$_2$ concentration than on a less weatherable planet. The relatively high concentrations of Ca and Mg (that ultimately sequester carbon through precipitation as carbonate) and the high weathering rates of mafic lithologies make it so that basaltic regions consume more CO$_2$ than regions where bedrock composition is closer to bulk continental crust \citep{Dessert2003a}. Data from basaltic watersheds show that chemical weathering rates are greatest in regions with high runoff, and as a result CO$_2$ consumption in basaltic regions is most pronounced in the tropical rain belt \citep{Dessert2003a, Hartmann2014a}.

@@

In the original models that proposed the `Fire and Ice' hypothesis as an explanation for the onset of the Sturtian `Snowball Earth' Glaciation, chemical weathering was modeled as an Arrhenius relationship where it was solely a function of temperature and run-off \citep{Donnadieu2004a}. However, such an approach neglects the effects that can result through soil shielding and regolith development in low-relief regions. Recent progress on understanding the relationships between landscapes, topography, and chemical weathering reveals that these effects are quite important (e.g. \citealp{Maher2014a}). As a result, more recent modeling of chemical weathering that has incorporated such processes highlights the importance of high-relief regions relative to low-relief ones \citep{Godderis2017b}. Without active uplift, soil shielding from regolith development on low-relief LIPs could significantly decrease the local weatherability of a LIP and mute its impact on global weatherability. In contrast, processes that lead to continued exhumation of mafic lithologies and the creation of steep topography, particularly in tropical regions, may exert a strong control on global weatherability and long-term climate. This interpretation underlies the hypothesis that arc-continent collisons in the tropics during the Ordovician \citep{Swanson-Hysell2017a} and the Cenozoic \citep{Jagoutz2016a} played a significant role in transitions into glacial climate states at those times.

A complication with the interpretation of soil shielding and limited weathering of LIPs is the rapid area decay rate inferred from the comparison of current LIP extent to estimated original extent. A couple considerations are relevant with respect to this analysis: 1) the current extent of LIP exposure is reduced in part by volcanics being covered by unconsolidated sediments (i.e. regolith development itself) in a number of the provinces; 2) the current extent of LIP exposure may be incomplete for some of the provinces; 3) the initial LIP extents are typically poorly constrained and are likely overestimates which could be resulting in inflated interpreted decay rates. Future efforts to develop better-constrained estimates of original LIP extent will improve analyses such as that in this contribution.

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$500~Myr. Furthermore, the approach of seeking to evaluate correlation between LIP area and glaciation is complicated for Neoproterozoic Snowball Earth events, since ice-albedo runaway leads to persistent global glaciation absent of continued forcing through normal carbon cycle processes until the build-up of sufficient pCO$_2$ to drive deglaciation. Nevertheless, the  hypothesis of tropical LIP area associated with the ca. 720~Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth Glaciation is a major motivating driver behind conducting this analysis. How does the tropical LIP area associated with the Franklin LIP compare to that observed in the Phanerozoic? Using the paleomagnetic pole of \citet{Denyszyn2009a}, we reconstruct the paleolatitude of the Franklin LIP at the time of emplacement, and find that $\sim$99.7\% (or $\sim$2.6~Mkm$^{2}$) of the LIP erupted within 15\textdegree\xspace of the equator. This Franklin LIP tropical area at the time of emplacement is approximately equivalent to the minor Cenozoic peak, and is significantly smaller than the other Phanerozoic peaks (Fig. \ref{fig:LIP_area}C). The ca. 1109~Ma Umkondo is another Precambrian LIP that is constrained to have erupted in the tropics, although it is not known to be associated with any glaciation. We reconstruct the paleolatitude of the Umkondo LIP at the time of emplacement using the paleomagnetic pole of \citet{Swanson-Hysell2015b}, and find that effectively all of the LIP (or $\sim$2.0~Mkm$^{2}$) erupted within the tropics, an area that is slightly smaller than that for the Franklin LIP (Fig. \ref{fig:LIP_area}C). Together, these results indicate that the tropical area of the Franklin LIP, when compared that of Phanerozoic as well as other Precambrian LIPs, is not particularly notable, and therefore that the increase in global weatherability associated with the emplacement of the Franklin LIP in the tropics is also not uniquely large, especially when considering the severity of the glaciation that followed it. This conclusion is consistent with the results of our Phanerozoic analysis, which suggests that LIP area and its associated impact on global weatherability is not the fundamental control on Earth's long-term climate state.

@@

\begin{table}[h!]
\scriptsize
\caption{place holder}
\noindent\makebox[\textwidth]{
\begin{tabular}{lrlrrlrrl}
\toprule
                name &    age &                     age ref. &  original area Mkm2 &  present area Mkm2 &                                 present area ref. &  present/original &  half-life &   buried \\
\midrule
      Columbia River &   16.0 &   Kasbohm and Schoene (2018) &                0.68 &               0.38 &                           Buchan and Ernst (2004) &              0.56 &       19.1 &       no \\
                Afar &   30.0 &  Courtillot and Renne (2003) &                2.05 &               0.63 &                              Coffin et al. (2006) &              0.31 &       17.8 &  partial \\
                NAIP &   62.0 &         Larsen et al. (2016) &                1.07 &               0.29 &     Buchan and Ernst (2004), Coffin et al. (2006) &              0.27 &       32.8 &  partial \\
              Deccan &   66.0 &               Schoene (2015) &                0.83 &               0.56 &                              Coffin et al. (2006) &              0.68 &      118.6 &       no \\
          Seychelles &   66.0 &               Schoene (2015) &                0.46 &               0.00 &                              Coffin et al. (2006) &              0.00 &        0.0 &      yes \\
          Madagascar &   90.0 &    Cucciniello et al. (2010) &                0.63 &               0.03 &                              Coffin et al. (2006) &              0.05 &       20.8 &       no \\
 Caribbean-Colombian &   94.0 &         Loewen et al. (2013) &                0.71 &               0.13 &                              Coffin et al. (2006) &              0.18 &       38.0 &       no \\
               HALIP &   95.0 &      Kingsbury et al. (2018) &                3.60 &               0.15 &                      Moosdorf and Hartmann (2015) &              0.04 &       20.5 &       no \\
              EQUAMP &  131.0 &       Hollanda et al. (2016) &                0.66 &               0.01 &                            Hollanda et al. (2016) &              0.01 &       19.7 &       no \\
               Comei &  132.0 &            Zhu et al. (2009) &                0.11 &                NaN &                                               n/a &               NaN &        NaN &       no \\
             Bunbury &  132.0 &            Zhu et al. (2009) &                0.03 &               0.00 &                              Thorne et al. (2014) &              0.05 &       30.5 &       no \\
     Parana-Etendeka &  135.0 &      Florisbal et al. (2014) &                3.12 &               0.40 &  Coffin et al. (2016), Teixeira and Albrez (2017) &              0.13 &       45.9 &  partial \\
                Trap &  140.0 &          Ernst et al. (2001) &                0.03 &               0.00 &                           Ernst and Buchan (2001) &              0.00 &        0.0 &       no \\
 NW Australia Margin &  160.0 &            COME BACK TO THIS &                0.62 &               0.00 &                              Coffin et al. (2006) &              0.00 &        0.0 &      yes \\
               Karoo &  183.0 &        Burgess et al. (2015) &                3.21 &               0.15 &                               de Kock compilation &              0.05 &       42.3 &       no \\
              Ferrar &  183.0 &        Burgess et al. (2015) &                0.18 &                NaN &                                               n/a &               NaN &        NaN &       no \\
                CAMP &  201.0 &      Blackburn et al. (2013) &               11.46 &               0.23 &                   Marzoli and Parisio compilation &              0.02 &       35.6 &  partial \\
             Siberia &  252.0 &   Burgess and Bowring (2015) &                3.46 &               0.47 &                              Coffin et al. (2006) &              0.14 &       88.8 &       no \\
            Emeishan &  259.0 &           Zhou et al. (2002) &                0.71 &               0.06 &                              Coffin et al. (2006) &              0.09 &       74.6 &       no \\
    Panjal-Qiangtang &  283.0 &           Zhai et al. (2013) &                0.11 &                NaN &                                               n/a &               NaN &        NaN &       no \\
               Tarim &  290.0 &             Xu et al. (2014) &                0.35 &                NaN &                                               n/a &               NaN &        NaN &       no \\
            Magdalen &  360.0 &         Murphy et al. (1999) &                0.42 &                NaN &                                               n/a &               NaN &        NaN &       no \\
              Vilyui &  374.0 &          Renne et al. (2010) &                1.14 &                NaN &                                               n/a &               NaN &        NaN &       no \\
        Kola-Dnieper &  380.0 &    Arzamastsev and Wu (2014) &                5.90 &                NaN &                                               n/a &               NaN &        NaN &       no \\
            Suordakh &  450.0 &       Khudoley et al. (2013) &                0.02 &                NaN &                                               n/a &               NaN &        NaN &       no \\
         Kalkarindji &  511.0 &        Jourdan et al. (2014) &                3.54 &               0.17 &                              Thorne et al. (2014) &              0.05 &      118.2 &       no \\
            Franklin &  720.0 &       Denyszyn et al. (2009) &                2.62 &               0.04 &                           Buchan and Ernst (2004) &              0.02 &      127.6 &       no \\
\bottomrule
\end{tabular}
}
\begin{tablenotes}[para,flushleft]
place holder.
\end{tablenotes}
\label{tab:LIPs}
\end{table}
